---
title: "02-25_932Hw1"
author: "Erika Sanchez"
date: "2/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Version-controlled directory system
   https://github.com/esanchezb/esb-agro932.git
 

# Download Phaseolus reference genome
#### from https://plants.ensembl.org/Phaseolus_vulgaris/Info/Index
```{bash, eval=FALSE}
wget ftp://ftp.ensemblgenomes.org/pub/plants/release-46/fasta/phaseolus_vulgaris/dna/Phaseolus_vulgaris.PhaVulg1_0.dna_sm.nonchromosomal.fa.gz
```

#### Unzip reference genome file
```{r}
gunzip Phaseolus_vulgaris.PhaVulg1_0.dna_sm.nonchromosomal.fa.gz 
```


# 2. Simulation strategy and the hypothesis to test (positive, negative, or neutral selection)

- Reference (about 6.268 kb)
    Phaseolus_vulgaris.v1_0.fa
- 20x coverage
    N 5000
- PE 100bp
    -1 100 -2 100
- Only SNP no Indel
    -R 0 -X 0
-	Mutation rate is low
    -r 0.01

#### simulation
```{bash, eval=FALSE}
for i in {1..10}; do    wgsim ../../largedata/Phaseolus_vulgaris.v1_0.fa -e 0 -d 500 -N 5000 -1 100 -2 100 -r 0.01  -R 0 -X 0 l$i.read1.fq l$i.read2.fq; done
```

#### count reads
```{r}
wc -l l7.read2.fq
```


# Align the NGS reads to the reference genome

```{bash, eval=FALSE}
module load bwa samtools

# index the reference genome
bwa index Phaseolus_vulgaris.v1_0.fa

# bwa mem to align the reads to the reference genome 
for i in {1..10}; do bwa mem ../../largedata/Phaseolus_vulgaris.v1_0.fa l$i.read1.fq l$i.read2.fq | samtools view -bSh - > l$i.bam; done

# sort 
for i in *.bam; do samtools sort $i -o sorted_$i; done

# index ("Indexing a genome sorted BAM file allows one to quickly extract alignments overlapping particular genomic regions")
for i in sorted*.bam; do samtools index $i; done

# samtools flagstat – counts the number of alignments for each FLAG type
samtools flagstat sorted_l1.bam
```

# A procedure to calculate $\theta$ and $F_{ST}$ values
```{r}

```



## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
