---
title: "02-25_932Hw1"
author: "Erika Sanchez"
date: "2/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Version-controlled directory system
   https://github.com/esanchezb/esb-agro932.git
 
# Download Phaseolus reference genome
#### from https://plants.ensembl.org/Phaseolus_vulgaris/Info/Index
```{bash, eval=FALSE}
wget ftp://ftp.ensemblgenomes.org/pub/plants/release-46/fasta/phaseolus_vulgaris/dna/Phaseolus_vulgaris.PhaVulg1_0.dna_sm.nonchromosomal.fa.gz
```

#### Unzip reference genome file
```{bash, eval=FALSE}
gunzip Phaseolus_vulgaris.PhaVulg1_0.dna_sm.nonchromosomal.fa.gz 

mv Phaseolus_vulgaris.PhaVulg1_0.dna_sm.nonchromosomal.fa PhaVulg1_0.fa

bwa index PhaVulg1_0.fa
```

# 2. Simulation strategy and the hypothesis to test (positive, negative, or neutral selection)

- Reference (about 6.268 kb)
    PhaVulg1_0.fa
- 20x coverage
    N 5000
- PE 100bp
    -1 100 -2 100
- Only SNP no Indel
    -R 0 -X 0
-	Mutation rate is low
    -r 0.01

#### simulation
```{bash, eval=FALSE}
for i in {1..10}; do    wgsim ../../largedata/PhaVulg1_0.fa -e 0 -d 500 -N 5000 -1 100 -2 100 -r 0.01  -R 0 -X 0 l$i.read1.fq l$i.read2.fq; done
```

#### count reads
```{bash, eval=FALSE}
wc -l l7.read2.fq
```

# Align the NGS reads to the reference genome

```{bash, eval=FALSE}
module load bwa samtools

# index the reference genome
bwa index Phaseolus_vulgaris.v1_0.fa

# bwa mem to align the reads to the reference genome 
for i in {1..10}; do bwa mem ../../largedata/PhaVulg1_0.fa l$i.read1.fq l$i.read2.fq | samtools view -bSh - > l$i.bam; done

# sort 
for i in *.bam; do samtools sort $i -o sorted_$i; done

# index ("Indexing a genome sorted BAM file allows one to quickly extract alignments overlapping particular genomic regions")
for i in sorted*.bam; do samtools index $i; done

# samtools flagstat – counts the number of alignments for each FLAG type
samtools flagstat sorted_l1.bam
```

```{bash, eval=FALSE}
# write the bam files to a txt file
mkdir bam_files
mv sorted*.bam bam_files
cd bam_files
ls sorted*.bam > bam.txt
```

```{bash, eval=FALSE}
# run angsd
../../../bin/angsd/angsd -bam bam.txt -doSaf 1 -anc ../../../largedata/PhaVulg1_0.fa -GL 1  -out out_ang

../../../bin/angsd/misc/realSFS out_ang.saf.idx > out.sfs

cp out.sfs ../../cache/
```

# Calculate $\theta$ values
```{bash, eval=FALSE}
../../../bin/angsd/angsd -bam bam.txt -out out_ang -doThetas 1 -doSaf 1 -pest out.sfs -anc ../../../largedata/PhaVulg1_0.fa -GL 1

../../../bin/angsd/misc/thetaStat print out_ang.thetas.idx > theta.txt

cp theta.txt ../../cache/
```

# Calculate $F_{ST}$ values
```{bash, eval=FALSE}

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
